---
title: "Bacteria 2021 Sterile Sentinels"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, )
knitr::opts_knit$set(root.dir = "/Users/sonya.erlandson/OneDrive - USDA/files to save/sterile sentinels/sr analysis")
```

## Load data
```{r}
comm <- read.csv("data tables/emu_16S_counts_2021.csv", header = TRUE, row.names = 1, check.names = F)
tax <- read.csv("data tables/emu_16S_taxonomy_2021.csv")
samples <- read.csv("data tables/16S_sample_data_2021_subset.csv")
taxtrait <- read.table("data tables/16s_taxonomy_with_traits_2021.tsv", sep = "\t", header = TRUE)
fapro <- read.table("FAPROTAX_DONE/functional_table_2021.tsv", sep = "\t", header = TRUE, row.names = 1)
leco <- readxl::read_xlsx("data tables/Alt Rotation Sentinel Soil 2020-2022.xlsx")

leco <- leco[leco$Year ==2021, ]
```

# Load R libraries and custom functions
```{r}
library(vegan)
library(tidyverse)
library(knitr)
library(ALDEx2)


source("r scripts/subset_samples_community_table.R")
source("r scripts/permANOVA_functions.R")

aggregateTaxonomy <- function(community, taxonomy, agglevel){
  commtranspose <- as.data.frame(t(community))
  whichlevel <- which(colnames(tax) == agglevel)
  commagg <- aggregate(commtranspose, by = list(tax[, whichlevel]), FUN = sum)
  # rowSums(commagg[,-1])
  # commagg <- commagg[-1, ]
  row.names(commagg) <- commagg$Group.1 # new row names of genera
  commagg <- commagg[, -1] # remove grouping variable
  commagg <- as.data.frame(t(commagg))
}

#subset community, samples, taxonomy. need subset r script
rotation_ttests <- function(comm, samples, type=NULL, time=NULL, test_name){
  test <- comm_subset(comm, samples, type, time) # subset community
  testsamp <- sample_subset(samples, type, time) # subset samples
  # aldex t-test
  aldextest <- ALDEx2::aldex(reads = t(test), conditions = testsamp$rotation, test = "t", effect = TRUE, mc.samples = 128)
  # add sample data and test information
  aldextest$type <- rep(test_name, length(aldextest$rab.all))
  aldextest$species <- row.names(aldextest)
  return(aldextest)
  
}

type_ttest <- function(comm, samples, time=NULL, test_name){
  test <- comm_subset(comm, samples, which_time =  time) # subset community
  testsamp <- sample_subset(samples,  which_time = time) # subset samples
  # aldex t-test
  aldextest <- ALDEx2::aldex(reads = t(test), conditions = testsamp$type, test = "t", effect = TRUE, mc.samples = 128)
  # add sample data and test information
  aldextest$type <- rep(test_name, length(aldextest$rab.all))
  aldextest$species <- row.names(aldextest)
  return(aldextest)
  
}

```


#Soil chemistry and nutrients
permanova? 
WOULD be better to combine 2020 and 2021 for this analysis.
```{r}
# need to aggregate community or expand soil data
bulk <- comm_subset(comm, samples, "bulk")
bulksamp <- sample_subset(samples, "bulk")

bulk <- bulk[order(bulksamp$plot), ]
bulksamp <- bulksamp[order(bulksamp$plot), ]

bulk0 <- bulk[bulksamp$time==0, ]

lecorep <- leco[match(bulksamp$plot, leco$Plot), ]

# need to standardize the soil variables.
lecostd <- leco[ -4,c(7,8,11:18)] # missig 220
lecostd <- data.frame(scale(lecostd))

soilpermanova <- adonis2(decostand(bulk0+1, method = "clr")~PctN+PctC+pH+SWC_g.drygsoil+NH4N_mg.kg+P.Bray_mg.kg, data=lecostd, method = "euclid", by = "margin")

soilpermanova


soilrda <- rda(decostand(bulk0+1, method = "clr")~PctN+PctC+pH+SWC_g.drygsoil+NH4N_mg.kg+P.Bray_mg.kg, data=lecostd, method = "euclid")
plot(soilrda)
anova(soilrda)


soilmoisturerda <- adonis2(decostand(comm+1, method = "clr")~samples$percent_wc+samples$type +samples$rotation, method = "euclid")
soilmoisturerda
```

# Statistical power
adonis r2 - cropping system at each time point
power of sentinels vs bulk and incubation time will be determined in another script with all four datasets
```{r, echo=FALSE}
bulkanova <- adonis2(decostand(comm_subset(comm+1, samples, "bulk"), method = "clr") ~ week + rotation, data = sample_subset(samples, "bulk"), method = "euclid", by = "terms")

baganova <- adonis2(decostand(comm_subset(comm+1, samples, "bag"), method = "clr")  ~ week + rotation, data = sample_subset(samples, "bag"), method = "euclid", by = "terms")

# for each time point:
bulkpermanovas <- bulkANOVAs(comm, samples)
bagpermanovas <- bagANOVAs(comm, samples)

allpermanovas <- rbind(bulkpermanovas, bagpermanovas)

write.csv(allpermanovas, "bacteria_2021_adonis_results.csv")
```

total dna, will be determined in another script with all four datasets
diversity too.
#Differential Abundance
###sentinels vs bulk soil
```{r, message=F, include=F}
allweeks <- rbind(type_ttest(comm, samples, time = 1, "week_1"),
                  type_ttest(comm, samples, time = 2, "week_2"),
                  type_ttest(comm, samples, time = 3, "week_3"),
                  type_ttest(comm, samples, time = 4, "week_4"),
                  type_ttest(comm, samples, time = 5, "week_5"))


allweeks$amplicon <- rep ("bacteria_2021", length(allweeks$rab.all))


```


##cropping system
####sentinels
```{r, message=F, include=F}
# aggregate taxonomic level
summary(colnames(comm) == tax$taxon_name_sh)
genuscomm <- aggregateTaxonomy(comm, tax, "genus")

bag <- rbind(rotation_ttests(comm, samples, type = "bag", time = 1, "bag_1"),
              rotation_ttests(comm, samples, type = "bag", time = 2, "bag_2"),
              rotation_ttests(comm, samples, type = "bag", time = 3, "bag_3"),
              rotation_ttests(comm, samples, type = "bag", time = 4, "bag_4"),
              rotation_ttests(comm, samples, type = "bag", time = 5, "bag_5"))


genusbag <- rbind(rotation_ttests(genuscomm, samples, type = "bag", time = 1, "bag_1"),
              rotation_ttests(genuscomm, samples, type = "bag", time = 2, "bag_2"),
              rotation_ttests(genuscomm, samples, type = "bag", time = 3, "bag_3"),
              rotation_ttests(genuscomm, samples, type = "bag", time = 4, "bag_4"),
              rotation_ttests(genuscomm, samples, type = "bag", time = 5, "bag_5"))


```

###bulk soil
```{r, echo=FALSE,message=F, include=F}

bulk <- rbind(rotation_ttests(comm, samples, type = "bulk", time = 0, "bulk_0"),
              rotation_ttests(comm, samples, type = "bulk", time = 1, "bulk_1"),
              rotation_ttests(comm, samples, type = "bulk", time = 2, "bulk_2"),
              rotation_ttests(comm, samples, type = "bulk", time = 3, "bulk_3"),
              rotation_ttests(comm, samples, type = "bulk", time = 4, "bulk_4"),
              rotation_ttests(comm, samples, type = "bulk", time = 5, "bulk_5"))


genusbulk <- rbind(rotation_ttests(genuscomm, samples, type = "bulk", time = 0, "bulk_0"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 1, "bulk_1"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 2, "bulk_2"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 3, "bulk_3"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 4, "bulk_4"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 5, "bulk_5"))


```

# diff abund summary
```{r}
rotationtests <- rbind(bag,bulk)
rotationfilt <- rotationtests[abs(rotationtests$effect)>1, ]
length(unique(rotationfilt$species))

genusrotationtests <- rbind(genusbag,genusbulk)
traittests <- rbind(bag.traits, bulk.traits)
# assign guilds
rotationfilt$primary_lifestyle <- tax$primary_lifestyle[match(rotationfilt$species, tax$taxon_name_sh)]
rotationfilt$secondary_lifestyle <- tax$secondary_lifestyle[match(rotationfilt$species, tax$taxon_name_sh)]

cstaxa <- rotationfilt[rotationfilt$effect < 0, ]
csswptaxa <- rotationfilt[rotationfilt$effect > 0, ]


bagtraits <- allweeks.traits[allweeks.traits$effect<-1, ]
bulktraits <- allweeks.traits[allweeks.traits$effect>1, ]

write.csv(rotationtests, "bacteria_2021_rotation_ttests_species.csv", quote = F)
write.csv(genusrotationtests, "bacteria_2021_rotation_ttests_genus.csv", quote = F)
write.csv(allweeks, "bacteria_2021_type_ttests_species.csv", quote = F)
#write.csv(allweeks.traits, "bacteria_2021_type_traits.csv", quote = F)
#write.csv(traittests, "bacteria_2021_rotation_traits.csv", quote = F)
```

#Diversity and Richness
```{r}
library(hillR)
#stats
summary(row.names(comm)==samples$fungi_barcode)
samples$week_int <- samples$week
samples$lib_size <- rowSums(comm)

calculate_diversity <- function(comm, samples){
  shannon <- diversity(comm, "shannon")
  hillq0 <- hill_taxa(comm, q=0, MARGIN = 1)
  # get rarefied estimates
  rar <- rrarefy(comm, sample = min(rowSums(comm)))
  rar.alpha <- specnumber(rar)
  rar.shannon <- diversity(rar, "shannon")
  rar.hillq0 <- hill_taxa(rar, q=0, MARGIN = 1)
  rar.hillq1 <- hill_taxa(rar, q=1, MARGIN = 1)
  rar.hillq2 <- hill_taxa(rar, q=2, MARGIN = 1)
  #add to sample data
  samples$shannon <- shannon
  samples$hill <- hillq0
  samples$rar.alpha <- rar.alpha
  samples$rar.shannon <- rar.shannon
  samples$rar.hillq0 <- rar.hillq0
  samples$rar.hillq1 <- rar.hillq1
  samples$rar.hillq2 <- rar.hillq2
  return(samples)
}

div.df <- calculate_diversity(comm, samples)

bagdiv <- div.df[div.df$type == "bag", ]
bulkdiv <- div.df[div.df$type == "bulk", ]

weeksubset0 <- bulkdiv[bulkdiv$time == 0, ]
t.test(rar.hillq1 ~ rotation, data = weeksubset0)
# no sig diff for bulk 0
alphadivtests <- list()
for(t in 1:5){
  print(paste("testing week", t))
  weeksubset <- bagdiv[bagdiv$time == t, ]
  test.res <- t.test(rar.hillq1 ~ rotation, data = weeksubset)
  alphadivtests[[t]] <- test.res
}

alphadivtests[[1]]
alphadivtests[[2]]
alphadivtests[[3]]
alphadivtests[[4]]
alphadivtests[[5]]


#!! warning !!! no longer matches up with community table
bagdiv <- bagdiv[order(bagdiv$plot), ]
bulkdiv <- bulkdiv[order(bulkdiv$plot), ]

ggplot(bulkdiv)+
  geom_point(aes(x = week_int, y = lib_size, color = rotation), size = 2)+
  theme_minimal()+
  scale_color_manual(values = c("skyblue", "hotpink"))+
  labs(y = "DNA concentration (ng.uL)", x = "week sampled",title = "bulk soil")
  stat_smooth(aes(x = week_int, y = rar.hillq1), method="lm", se=TRUE, fill=NA,
              formula=y ~ x ,colour="gray")
  
ggplot(bagdiv)+
  geom_point(aes(x = week_int, y = lib_size, color = rotation), size = 3)+
  theme_minimal()+
  scale_color_manual(values = c("skyblue", "hotpink"))+
  labs(y = "DNA concentration (ng.uL)",x = "week sampled", title = "bag soil")+
  stat_smooth(aes(x = week_int, y = DNA_ng.ul), method="lm", se=TRUE, fill=NA,
              formula=y ~ exp(-x), colour="gray")


t.test(bagdiv$rar.alpha ~ as.numeric(as.factor(bagdiv$rotation)))
t.test(bagdiv$rar.hillq1 ~ as.numeric(as.factor(bagdiv$rotation)))
t.test(bagdiv$rar.shannon ~ as.numeric(as.factor(bagdiv$rotation)))

t.test(bulkdiv$rar.alpha ~ as.numeric(as.factor(bulkdiv$rotation)))
t.test(bulkdiv$rar.hillq1 ~ as.numeric(as.factor(bulkdiv$rotation)))
t.test(bulkdiv$rar.shannon ~ as.numeric(as.factor(bulkdiv$rotation)))



mean(bagdiv$rar.alpha[bagdiv$rotation == "CS"])
mean(bagdiv$rar.alpha[bagdiv$rotation == "CSSwP"])

sd(bagdiv$rar.alpha[bagdiv$rotation == "CS"])
sd(bagdiv$rar.alpha[bagdiv$rotation == "CSSwP"])

mean(bagdiv$rar.hillq1[bagdiv$rotation == "CS"])
mean(bagdiv$rar.hillq1[bagdiv$rotation == "CSSwP"])

sd(bagdiv$rar.hillq1[bagdiv$rotation == "CS"])
sd(bagdiv$rar.hillq1[bagdiv$rotation == "CSSwP"])

#####################################

mean(bulkdiv$rar.alpha[bulkdiv$rotation == "CS"])
mean(bulkdiv$rar.alpha[bulkdiv$rotation == "CSSwP"])

sd(bulkdiv$rar.alpha[bulkdiv$rotation == "CS"])
sd(bulkdiv$rar.alpha[bulkdiv$rotation == "CSSwP"])


mean(bulkdiv$rar.hillq1[bulkdiv$rotation == "CS"])
mean(bulkdiv$rar.hillq1[bulkdiv$rotation == "CSSwP"])

sd(bulkdiv$rar.hillq1[bulkdiv$rotation == "CS"])
sd(bulkdiv$rar.hillq1[bulkdiv$rotation == "CSSwP"])

```


# Overall RDA
```{r}
commclr <- decostand(comm+1, method = "clr")
allrda <- rda(commclr ~ type + week + rotation, data = samples)
allrda
summary(allrda)
anova(allrda, by = "terms")
anova(allrda, by = "axis")

source("r scripts/RDA outlier functions.R")
adaptive <- rdadapt(allrda, 3) # finds taxa that contribute most to each axis, returns a list of pvalues

# get taxa that contribute most to each axis
taxadf <- scores(allrda, scaling = 2)$species
taxadf <- cbind(taxadf, adaptive)
taxadf$spplabels <- sapply(strsplit(rownames(taxadf), " "), "[[", 1)
# filter to just smallest qvalues
taxadf.filt <- taxadf[order(taxadf$q.values),]
taxadf.filt <- taxadf.filt[1:20,]


rdaplot <- ggplot()+
    # site scores
    geom_point(aes(scores(allrda, scaling = 2)$sites[,1], 
                   scores(allrda, scaling = 2)$sites[,2], 
                   shape = samples$rotation, 
                   color = as.factor(samples$week)), 
               size = 4)+
  # specis scores  
  # geom_point(aes(scores(allrda, scaling = 2)$species[,1], 
  #                    scores(allrda, scaling = 2)$species[,2]), 
  #                alpha = 1/10, color = "gray25", size = 1)+
  # species labels
    # geom_text_repel(aes(x = taxadf.filt[ ,1],
    #               y = taxadf.filt[ ,2],
    #               label =taxadf.filt$spplabels), max.overlaps = 30,
    #           size = 3, color = "gray25", fontface = "italic")+
    geom_text(aes(x = scores(allrda, scaling = 2)$centroids[ ,1]-0.3,
                  y = scores(allrda, scaling = 2)$centroids[ ,2],
                  label = c("sentinel", "bulk", "CS", "CSSwP")), check_overlap = FALSE,
              size = 4, fontface = "bold")+
    xlab("RDA1 - 53% of total variance, p = 0.001")+
    ylab("RDA2 - 3.2% of total variance, p = 0.001")+
    labs(color = "week", shape = "crop\nrotation")+
    theme(legend.text = element_text(size = 12, face = "italic"))+
    theme_minimal()
rdaplot

ggsave(filename = "bac_2021_allRDA_scale2.tiff", plot = rdaplot, device = "tiff", width = 6, height = 4, units = "in")
```


# Sample Type RDAs
```{r}
bulk <- comm_subset(comm, samples, "bulk")
bulksamp <- sample_subset(samples, "bulk")

bulkclr <- decostand(bulk+1, method = "clr")
bulkrda <- rda(bulkclr ~  week + rotation, data = bulksamp)
bulkrda
summary(bulkrda)
anova(bulkrda, by = "terms")
anova(bulkrda, by = "axis")

source("r scripts/RDA outlier functions.R")
adaptive <- rdadapt(bulkrda, 2) # finds taxa that contribute most to each axis, returns a list of pvalues

# get taxa that contribute most to each axis
taxadf <- scores(bulkrda, scaling = 2)$species
taxadf <- cbind(taxadf, adaptive)
taxadf$spplabels <- sapply(strsplit(rownames(taxadf), " "), "[[", 1)
# filter to just smallest qvalues
taxadf.filt <- taxadf[order(taxadf$q.values),]
taxadf.filt <- taxadf.filt[1:20,]


bulkrdaplot <- ggplot()+
    # site scores
    geom_point(aes(scores(bulkrda, scaling = 2)$sites[,1], 
                   scores(bulkrda, scaling = 2)$sites[,2], 
                   shape = bulksamp$rotation, 
                   color = as.factor(bulksamp$week)), 
               size = 4)+
  # species scores  
  geom_point(aes(scores(bulkrda, scaling = 2)$species[,1], 
                     scores(bulkrda, scaling = 2)$species[,2]), 
                 alpha = 1/10, color = "gray25", size = 1)+
  # species labels
    geom_text_repel(aes(x = taxadf.filt[ ,1],
                  y = taxadf.filt[ ,2],
                  label =taxadf.filt$spplabels), max.overlaps = 40,
              size = 3, color = "gray25", fontface = "italic")+
    geom_text(aes(x = scores(bulkrda, scaling = 2)$centroids[ ,1],
                  y = scores(bulkrda, scaling = 2)$centroids[ ,2],
                  label = c("CS", "CSSwP")), check_overlap = FALSE,
              size = 4, fontface = "bold")+
    xlab("RDA1 - 4.6% of total variance, p = 0.001")+
    ylab("RDA2 - 2.2% of total variance, p > 0.05")+
    labs(color = "week", shape = "crop\nrotation")+
    theme(legend.text = element_text(size = 12, face = "italic"))+
    theme_minimal()
bulkrdaplot

ggsave(filename = "bac_2021_bulkrda_scale2.tiff", plot = bulkrdaplot, device = "tiff", width = 6, height = 4, units = "in")


######################################

bag <- comm_subset(comm, samples, "bag")
bagsamp <- sample_subset(samples, "bag")

bagclr <- decostand(bag+1, method = "clr")
bagrda <- rda(bagclr ~  week + rotation, data = bagsamp)
bagrda
summary(bagrda)
anova(bagrda, by = "terms")
anova(bagrda, by = "axis")

source("r scripts/RDA outlier functions.R")
adaptive <- rdadapt(bagrda, 2) # finds taxa that contribute most to each axis, returns a list of pvalues

# get taxa that contribute most to each axis
taxadf <- scores(bagrda, scaling = 2)$species
taxadf <- cbind(taxadf, adaptive)
taxadf$spplabels <- sapply(strsplit(rownames(taxadf), " "), "[[", 1)
# filter to just smallest qvalues
taxadf.filt <- taxadf[order(taxadf$q.values),]
taxadf.filt <- taxadf.filt[1:20,]


bagrdaplot <- ggplot()+
    # site scores
    geom_point(aes(scores(bagrda, scaling = 2)$sites[,1], 
                   scores(bagrda, scaling = 2)$sites[,2], 
                   shape = bagsamp$rotation, 
                   color = as.factor(bagsamp$week)), 
               size = 4)+
  # species scores  
  geom_point(aes(scores(bagrda, scaling = 2)$species[,1], 
                     scores(bagrda, scaling = 2)$species[,2]), 
                 alpha = 1/10, color = "gray25", size = 1)+
  # species labels
    geom_text_repel(aes(x = taxadf.filt[ ,1],
                  y = taxadf.filt[ ,2],
                  label =taxadf.filt$spplabels), max.overlaps = 40,
              size = 3, color = "gray25", fontface = "italic")+
    geom_text(aes(x = scores(bagrda, scaling = 2)$centroids[ ,1],
                  y = scores(bagrda, scaling = 2)$centroids[ ,2],
                  label = c("CS", "CSSwP")), check_overlap = FALSE,
              size = 4, fontface = "bold")+
    xlab("RDA1 - 24% of total variance, p = 0.001")+
    ylab("RDA2 - 4.6% of total variance, p = 0.001")+
    labs(color = "week", shape = "crop\nrotation")+
    theme(legend.text = element_text(size = 12, face = "italic"))+
    theme_minimal()
bagrdaplot

ggsave(filename = "bac_2021_bagrda_scale2.tiff", plot = bagrdaplot, device = "tiff", width = 6, height = 4, units = "in")
```