---
title: "Bacteria 2020 Sterile Sentinels"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, )
knitr::opts_knit$set(root.dir = "/Users/sonya.erlandson/OneDrive - USDA/files to save/sterile sentinels/sr analysis")
```

## Load data
```{r}
# 2020 Bacteria
comm <- read.csv("data tables/emu_16S_counts_2020.csv", header = TRUE, row.names = 1, check.names = F)
tax <- read.csv("data tables/emu_16S_taxonomy_2020.csv", header = TRUE)
samples <- read.csv("data tables/16S_sample_data_2020.csv")
taxtrait <- read.table("data tables/16s_taxonomy_with_traits_2020.tsv", sep = "\t", header = TRUE)
fapro <- read.table("FAPROTAX_DONE/functional_table_2020.tsv", sep = "\t", header = TRUE, row.names = 1)
traitcomm <- as.data.frame(t(fapro))
leco <- readxl::read_xlsx("data tables/Alt Rotation Sentinel Soil 2020-2022.xlsx")

leco <- leco[leco$Year ==2020, ]
```

# Load R libraries and custom functions
```{r}
library(vegan)
library(tidyverse)
library(knitr)
library(ALDEx2)
library(betapart)

source("r scripts/subset_samples_community_table.R")
source("r scripts/permANOVA_functions.R")

aggregateTaxonomy <- function(community, taxonomy, agglevel){
  commtranspose <- as.data.frame(t(community))
  whichlevel <- which(colnames(tax) == agglevel)
  commagg <- aggregate(commtranspose, by = list(tax[, whichlevel]), FUN = sum)
  # rowSums(commagg[,-1])
  # commagg <- commagg[-1, ]
  row.names(commagg) <- commagg$Group.1 # new row names of genera
  commagg <- commagg[, -1] # remove grouping variable
  commagg <- as.data.frame(t(commagg))
}

#subset community, samples, taxonomy. need subset r script
rotation_ttests <- function(comm, samples, type=NULL, time=NULL, test_name){
  test <- comm_subset(comm, samples, type, time) # subset community
  testsamp <- sample_subset(samples, type, time) # subset samples
  # aldex t-test
  aldextest <- ALDEx2::aldex(reads = t(test), conditions = testsamp$rotation, test = "t", effect = TRUE, mc.samples = 128)
  # add sample data and test information
  aldextest$type <- rep(test_name, length(aldextest$rab.all))
  aldextest$species <- row.names(aldextest)
  return(aldextest)
  
}

type_ttest <- function(comm, samples, time=NULL, test_name){
  test <- comm_subset(comm, samples, which_time =  time) # subset community
  testsamp <- sample_subset(samples,  which_time = time) # subset samples
  # aldex t-test
  aldextest <- ALDEx2::aldex(reads = t(test), conditions = testsamp$type, test = "t", effect = TRUE, mc.samples = 128)
  # add sample data and test information
  aldextest$type <- rep(test_name, length(aldextest$rab.all))
  aldextest$species <- row.names(aldextest)
  return(aldextest)
  
}

```


#Soil chemistry and nutrients
permanova? 
WOULD be better to combine 2020 and 2021 for this analysis.
```{r}

# need to aggregate community or expand soil data
bulk <- comm_subset(comm, samples, "bulk")
bulksamp <- sample_subset(samples, "bulk")

bulk <- bulk[order(bulksamp$plot), ]
bulksamp <- bulksamp[order(bulksamp$plot), ]

bulk0 <- bulk[bulksamp$time==0, ]

lecorep <- leco[match(bulksamp$plot, leco$Plot), ]

# need to standardize the soil variables.
lecostd <- leco[ ,c(7,8,11:18)]
lecostd <- data.frame(scale(lecostd))

soilpermanova <- adonis2(decostand(bulk[bulksamp$time==0, ]+1, method = "clr")~PctN+PctC+pH+SWC_g.drygsoil+NH4N_mg.kg+P.Bray_mg.kg, data=lecostd, method = "euclid", by = "margin")

soilpermanova


soilrda <- rda(decostand(bulk0+1, method = "clr")~PctN+PctC+pH+SWC_g.drygsoil+NH4N_mg.kg+P.Bray_mg.kg, data=lecostd, method = "euclid")
plot(soilrda)
anova(soilrda)


soilmoisturepermanova <- adonis2(decostand(comm+1, method = "clr")~samples$percent_wc+samples$type+
                                   samples$rotation, method = "euclid")
soilmoisturepermanova
```

# Statistical power
adonis r2 - cropping system at each time point
power of sentinels vs bulk and incubation time will be determined in another script with all four datasets
```{r, echo=FALSE}
bulkanova <- adonis2(decostand(comm_subset(comm+1, samples, "bulk"), method = "clr") ~ rotation, data = sample_subset(samples, "bulk"), method = "euclid", by = "terms")

baganova <- adonis2(decostand(comm_subset(comm+1, samples, "bag"), method = "clr")  ~ week + rotation, data = sample_subset(samples, "bag"), method = "euclid", by = "terms")

# for each time point:
bulkpermanovas <- bulkANOVAs(comm, samples)
bagpermanovas <- bagANOVAs(comm, samples)

allpermanovas <- rbind(bulkpermanovas, bagpermanovas)

write.csv(allpermanovas, "bacteria_2020_adonis_results.csv")
```

total dna, will be determined in another script with all four datasets
diversity too.
#Differential Abundance
###sentinels vs bulk soil
```{r, message=F, include=F}
allweeks <- rbind(type_ttest(comm, samples, time = 1, "week_1"),
                  type_ttest(comm, samples, time = 2, "week_2"),
                  type_ttest(comm, samples, time = 3, "week_3"),
                  type_ttest(comm, samples, time = 4, "week_4"),
                  type_ttest(comm, samples, time = 5, "week_5"))


allweeks$amplicon <- rep ("bacteria_2020", length(allweeks$rab.all))


```

##cropping system
####sentinels + traits
```{r, message=F, include=F}
# aggregate taxonomic level
summary(colnames(comm) == tax$species)
genuscomm <- aggregateTaxonomy(comm, tax, "genus")

bag <- rbind(rotation_ttests(comm, samples, type = "bag", time = 1, "bag_1"),
              rotation_ttests(comm, samples, type = "bag", time = 2, "bag_2"),
              rotation_ttests(comm, samples, type = "bag", time = 3, "bag_3"),
              rotation_ttests(comm, samples, type = "bag", time = 4, "bag_4"),
              rotation_ttests(comm, samples, type = "bag", time = 5, "bag_5"))


genusbag <- rbind(rotation_ttests(genuscomm, samples, type = "bag", time = 1, "bag_1"),
              rotation_ttests(genuscomm, samples, type = "bag", time = 2, "bag_2"),
              rotation_ttests(genuscomm, samples, type = "bag", time = 3, "bag_3"),
              rotation_ttests(genuscomm, samples, type = "bag", time = 4, "bag_4"),
              rotation_ttests(genuscomm, samples, type = "bag", time = 5, "bag_5"))


```

###bulk soil + traits
```{r, echo=FALSE,message=F, include=F}

bulk <- rbind(rotation_ttests(comm, samples, type = "bulk", time = 0, "bulk_0"),
              rotation_ttests(comm, samples, type = "bulk", time = 1, "bulk_1"),
              rotation_ttests(comm, samples, type = "bulk", time = 2, "bulk_2"),
              rotation_ttests(comm, samples, type = "bulk", time = 3, "bulk_3"),
              rotation_ttests(comm, samples, type = "bulk", time = 4, "bulk_4"),
              rotation_ttests(comm, samples, type = "bulk", time = 5, "bulk_5"))


genusbulk <- rbind(rotation_ttests(genuscomm, samples, type = "bulk", time = 0, "bulk_0"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 1, "bulk_1"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 2, "bulk_2"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 3, "bulk_3"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 4, "bulk_4"),
              rotation_ttests(genuscomm, samples, type = "bulk", time = 5, "bulk_5"))

```

# diff abund summary
```{r}
rotationtests <- rbind(bag,bulk)
rotationfilt <- rotationtests[abs(rotationtests$effect)>1, ]
length(unique(rotationfilt$species))

genusrotationtests <- rbind(genusbag,genusbulk)
traittests <- rbind(bag.traits, bulk.traits)
# assign guilds
rotationfilt$primary_lifestyle <- tax$primary_lifestyle[match(rotationfilt$species, tax$taxon_name_sh)]
rotationfilt$secondary_lifestyle <- tax$secondary_lifestyle[match(rotationfilt$species, tax$taxon_name_sh)]

cstaxa <- rotationfilt[rotationfilt$effect < 0, ]
csswptaxa <- rotationfilt[rotationfilt$effect > 0, ]


bagtraits <- allweeks.traits[allweeks.traits$effect<-1, ]
bulktraits <- allweeks.traits[allweeks.traits$effect>1, ]

write.csv(rotationtests, "bacteria_2020_rotation_ttests_species.csv", quote = F)
write.csv(genusrotationtests, "bacteria_2020_rotation_ttests_genus.csv", quote = F)
write.csv(allweeks, "bacteria_2020_type_ttests_species.csv", quote = F)
write.csv(allweeks.traits, "bacteria_2020_type_traits.csv", quote = F)
write.csv(traittests, "bacteria_2020_rotation_traits.csv", quote = F)
```

#Diversity and Richness
```{r}
library(hillR)
#stats
set.seed(44)
summary(row.names(comm)==samples$bacteria_barcode)
samples$week_int <- samples$week
samples$lib_size <- rowSums(comm)

calculate_diversity <- function(comm, samples){
  shannon <- diversity(comm, "shannon")
  hillq0 <- hill_taxa(comm, q=0, MARGIN = 1)
  # get rarefied estimates
  rar <- rrarefy(comm, sample = min(rowSums(comm)))
  rar.alpha <- specnumber(rar)
  rar.shannon <- diversity(rar, "shannon")
  rar.hillq0 <- hill_taxa(rar, q=0, MARGIN = 1)
  rar.hillq1 <- hill_taxa(rar, q=1, MARGIN = 1)
  rar.hillq2 <- hill_taxa(rar, q=2, MARGIN = 1)
  #add to sample data
  samples$shannon <- shannon
  samples$hill <- hillq0
  samples$rar.alpha <- rar.alpha
  samples$rar.shannon <- rar.shannon
  samples$rar.hillq0 <- rar.hillq0
  samples$rar.hillq1 <- rar.hillq1
  samples$rar.hillq2 <- rar.hillq2
  return(samples)
}

div.df <- calculate_diversity(comm, samples)

bagdiv <- div.df[div.df$type == "bag", ]
bulkdiv <- div.df[div.df$type == "bulk", ]

weeksubset0 <- bulkdiv[bulkdiv$time == 0, ]
t.test(rar.hillq1 ~ rotation, data = weeksubset0)
# no sig diff for bulk 0
alphadivtests <- list()
for(t in 1:5){
  print(paste("testing week", t))
  weeksubset <- bulkdiv[bulkdiv$time == t, ]
  test.res <- t.test(rar.alpha ~ rotation, data = weeksubset)
  alphadivtests[[t]] <- test.res
}

alphadivtests[[1]]
alphadivtests[[2]]
alphadivtests[[3]]
alphadivtests[[4]]
alphadivtests[[5]]

# sentinels diversity not sig diff.
# sentinel richness not sig diff
# bulk soil richness and diversity not sig diff

#!! warning !!! no longer matches up with community table
bagdiv <- bagdiv[order(bagdiv$plot), ]
bulkdiv <- bulkdiv[order(bulkdiv$plot), ]

ggplot(bulkdiv)+
  geom_point(aes(x = week_int, y = rar.alpha, color = rotation), size = 2)+
  theme_minimal()+
  scale_color_manual(values = c("skyblue", "hotpink"))+
  labs(y = "DNA concentration (ng.uL)", x = "week sampled",title = "bulk soil")
  stat_smooth(aes(x = week_int, y = rar.hillq1), method="lm", se=TRUE, fill=NA,
              formula=y ~ x ,colour="gray")

ggplot(bagdiv)+
  geom_point(aes(x = week_int, y = rar.alpha, color = rotation), size = 3)+
  theme_minimal()+
  scale_color_manual(values = c("skyblue", "hotpink"))+
  labs(y = "richness",x = "week sampled", title = "bag soil")+
  stat_smooth(aes(x = week_int, y = rar.alpha), method="lm", se=TRUE, fill=NA,
              formula=y ~ exp(-x), colour="gray")


t.test(bagdiv$rar.alpha ~ as.numeric(as.factor(bagdiv$rotation)))
t.test(bagdiv$rar.hillq1 ~ as.numeric(as.factor(bagdiv$rotation)))
t.test(bagdiv$rar.shannon ~ as.numeric(as.factor(bagdiv$rotation)))

t.test(bulkdiv$rar.alpha ~ as.numeric(as.factor(bulkdiv$rotation)))
t.test(bulkdiv$rar.hillq1 ~ as.numeric(as.factor(bulkdiv$rotation)))
t.test(bulkdiv$rar.shannon ~ as.numeric(as.factor(bulkdiv$rotation)))

mean(bagdiv$rar.alpha[bagdiv$rotation == "CS"])
mean(bagdiv$rar.alpha[bagdiv$rotation == "CSSwP"])

sd(bagdiv$rar.alpha[bagdiv$rotation == "CS"])
sd(bagdiv$rar.alpha[bagdiv$rotation == "CSSwP"])

mean(bagdiv$rar.hillq1[bagdiv$rotation == "CS"])
mean(bagdiv$rar.hillq1[bagdiv$rotation == "CSSwP"])

sd(bagdiv$rar.hillq1[bagdiv$rotation == "CS"])
sd(bagdiv$rar.hillq1[bagdiv$rotation == "CSSwP"])

#####################################

mean(bulkdiv$rar.alpha[bulkdiv$rotation == "CS"])
mean(bulkdiv$rar.alpha[bulkdiv$rotation == "CSSwP"])

sd(bulkdiv$rar.alpha[bulkdiv$rotation == "CS"])
sd(bulkdiv$rar.alpha[bulkdiv$rotation == "CSSwP"])


mean(bulkdiv$rar.hillq1[bulkdiv$rotation == "CS"])
mean(bulkdiv$rar.hillq1[bulkdiv$rotation == "CSSwP"])

sd(bulkdiv$rar.hillq1[bulkdiv$rotation == "CS"])
sd(bulkdiv$rar.hillq1[bulkdiv$rotation == "CSSwP"])

```

# Temporal Beta diversity Analysis

```{r}
library(codyn)
# start with all time 1 vs all time 2, t2-t3, t3-t4, t4-t5. 
commrar <- as.data.frame(rrarefy(comm, min(rowSums(comm))))

bag <- comm_subset(commrar, samples, samp_type = "bag")
bagsamp <- sample_subset(samples, samp_type = "bag")

# each data matrix need to be the same dimension
t1 <- bag[bagsamp$time == 1, ];row.names(t1) <- bagsamp$plot[bagsamp$time == 1]
t2 <- bag[bagsamp$time == 2, ];row.names(t2) <- bagsamp$plot[bagsamp$time == 2]
t3 <- bag[bagsamp$time == 3, ];row.names(t3) <- bagsamp$plot[bagsamp$time == 3]
t4 <- bag[bagsamp$time == 4, ];row.names(t4) <- bagsamp$plot[bagsamp$time == 4]
t5 <- bag[bagsamp$time == 5, ];row.names(t5) <- bagsamp$plot[bagsamp$time == 5]


#transform data to presence absence
t1 <- decostand(t1, method = "pa")
t2 <- decostand(t2, method = "pa")
t3 <- decostand(t3, method = "pa")
t4 <- decostand(t4, method = "pa")
t5 <- decostand(t5, method = "pa")

# calculate jaccard indices
t1.2 <- beta.temp(t1, t2, "jacc"); colnames(t1.2) <- c("beta.jtu.1.2","beta.jne.1.2","beta.jac.1.2")
t2.3 <- beta.temp(t2, t3, "jacc"); colnames(t2.3) <- c("beta.jtu.2.3","beta.jne.2.3","beta.jac.2.3")
t3.4 <- beta.temp(t3, t4, "jacc"); colnames(t3.4) <- c("beta.jtu.3.4","beta.jne.3.4","beta.jac.3.4")
t4.5 <- beta.temp(t4, t5, "jacc"); colnames(t4.5) <- c("beta.jtu.4.5","beta.jne.4.5","beta.jac.4.5")

#make a dataframe for plotting
jacc.turnover <- c(t1.2$beta.jtu.1.2, t2.3$beta.jtu.2.3, t3.4$beta.jtu.3.4, t4.5$beta.jtu.4.5)
jacc.nestedness <- c(t1.2$beta.jne.1.2, t2.3$beta.jne.2.3, t3.4$beta.jne.3.4, t4.5$beta.jne.4.5)
jacc.dissimilarity <- c(t1.2$beta.jac.1.2, t2.3$beta.jac.2.3, t3.4$beta.jac.3.4, t4.5$beta.jac.4.5)
plot_number <- row.names(t1)
sampletime <- c(rep("1-2", 8), rep("2-4", 8), rep("4-8", 8), rep("8-12", 8))

temporal.betadiver <- data.frame(rep(plot_number, 4), sampletime, jacc.turnover, jacc.nestedness, jacc.dissimilarity)
colnames(temporal.betadiver)[1] <- "plot_number"
long.temporal.betadiver <- tidyr::pivot_longer(temporal.betadiver, cols = c("jacc.dissimilarity", "jacc.turnover", "jacc.nestedness"))
colnames(long.temporal.betadiver) <- c("plot", "week_interval", "beta_diversity", "value")

dissim.only <- long.temporal.betadiver[long.temporal.betadiver$beta_diversity == "jacc.dissimilarity", ]
# plot all on one
ggplot(long.temporal.betadiver, aes(x = week_interval, 
                                    y = value, 
                                    color = beta_diversity, 
                                    #group = plot_number, 
                                    #shape = beta_diversity)
       ))+
  geom_point(size = 4)+
  xlab("Week Interval")+
  ylab(" Partitioned Beta Diversity")+
  ggtitle("sentinel bacterial community dissimilarity across successive intervals")+
  #scale_shape_manual(values = c(19 , 12 ,2))+
  theme(legend.title = element_blank())

dissim.only$interval <- c(rep(1,8), rep(2,8), rep(3,8), rep(4,8))
summary(lm(value~interval, data = dissim.only))
# "One is species replacement (i.e. turnover), which consists in the substitution of species in one site by different species in the other site."
# "The second way is species loss (or gain), which implies the elimination (or addition) of species in only one of the sites, and leads to the poorest assemblage being a strict subset of the richest one (a pattern called nestedness)."



## Bulk soil

commrar <- as.data.frame(rrarefy(comm, min(rowSums(comm))))

bulk <- comm_subset(commrar, samples, samp_type = "bulk")
bulksamp <- sample_subset(samples, samp_type = "bulk")

# each data matrix need to be the same dimension
t1 <- bulk[bulksamp$time == 1, ];row.names(t1) <- bulksamp$plot[bulksamp$time == 1]
t2 <- bulk[bulksamp$time == 2, ];row.names(t2) <- bulksamp$plot[bulksamp$time == 2]
t3 <- bulk[bulksamp$time == 3, ];row.names(t3) <- bulksamp$plot[bulksamp$time == 3]
t4 <- bulk[bulksamp$time == 4, ];row.names(t4) <- bulksamp$plot[bulksamp$time == 4]
t5 <- bulk[bulksamp$time == 5, ];row.names(t5) <- bulksamp$plot[bulksamp$time == 5]


#transform data to presence absence
t1 <- decostand(t1, method = "pa")
t2 <- decostand(t2, method = "pa")
t3 <- decostand(t3, method = "pa")
t4 <- decostand(t4, method = "pa")
t5 <- decostand(t5, method = "pa")

# calculate jaccard indices
t1.2 <- beta.temp(t1, t2, "jacc"); colnames(t1.2) <- c("beta.jtu.1.2","beta.jne.1.2","beta.jac.1.2")
t2.3 <- beta.temp(t2, t3, "jacc"); colnames(t2.3) <- c("beta.jtu.2.3","beta.jne.2.3","beta.jac.2.3")
t3.4 <- beta.temp(t3, t4, "jacc"); colnames(t3.4) <- c("beta.jtu.3.4","beta.jne.3.4","beta.jac.3.4")
t4.5 <- beta.temp(t4, t5, "jacc"); colnames(t4.5) <- c("beta.jtu.4.5","beta.jne.4.5","beta.jac.4.5")

#make a dataframe for plotting
jacc.turnover <- c(t1.2$beta.jtu.1.2, t2.3$beta.jtu.2.3, t3.4$beta.jtu.3.4, t4.5$beta.jtu.4.5)
jacc.nestedness <- c(t1.2$beta.jne.1.2, t2.3$beta.jne.2.3, t3.4$beta.jne.3.4, t4.5$beta.jne.4.5)
jacc.dissimilarity <- c(t1.2$beta.jac.1.2, t2.3$beta.jac.2.3, t3.4$beta.jac.3.4, t4.5$beta.jac.4.5)
plot_number <- row.names(t1)
sampletime <- c(rep("1-2", 8), rep("2-4", 8), rep("4-8", 8), rep("8-12", 8))

temporal.betadiver <- data.frame(rep(plot_number, 4), sampletime, jacc.turnover, jacc.nestedness, jacc.dissimilarity)
colnames(temporal.betadiver)[1] <- "plot_number"
long.temporal.betadiver <- tidyr::pivot_longer(temporal.betadiver, cols = c("jacc.dissimilarity", "jacc.turnover", "jacc.nestedness"))
colnames(long.temporal.betadiver) <- c("plot", "week_interval", "beta_diversity", "value")

dissim.only <- long.temporal.betadiver[long.temporal.betadiver$beta_diversity == "jacc.dissimilarity", ]
# plot all on one
ggplot(long.temporal.betadiver, aes(x = week_interval, 
                                    y = value, 
                                    color = beta_diversity, 
                                    #group = plot_number, 
                                    #shape = beta_diversity)
       ))+
  geom_point(size = 4)+
  xlab("Week Interval")+
  ylab(" Partitioned Beta Diversity")+
  ggtitle("bulk soil bacterial community dissimilarity across successive intervals")+
  #scale_shape_manual(values = c(19 , 12 ,2))+
  theme(legend.title = element_blank())

dissim.only$interval <- c(rep(1,8), rep(2,8), rep(3,8), rep(4,8))
summary(lm(value~interval, data = dissim.only))
# "One is species replacement (i.e. turnover), which consists in the substitution of species in one site by different species in the other site."
# "The second way is species loss (or gain), which implies the elimination (or addition) of species in only one of the sites, and leads to the poorest assemblage being a strict subset of the richest one (a pattern called nestedness)."

```

# Overall RDA
```{r}
commclr <- decostand(comm+1, method = "clr")
allrda <- rda(commclr ~ type + week + rotation, data = samples)
allrda
summary(allrda)
anova(allrda, by = "margin")
anova(allrda, by = "axis")

# get taxa to plot
source("r scripts/RDA outlier functions.R")
adaptive <- rdadapt(allrda, 3) # finds taxa that contribute most to each axis, returns a list of pvalues

# get taxa that contribute most to each axis
taxadf <- scores(allrda, scaling = 2)$species
taxadf <- cbind(taxadf, adaptive)
taxadf$spplabels <- sapply(strsplit(rownames(taxadf), " "), "[[", 1)
# filter to just smallest qvalues
taxadf.filt <- taxadf[order(taxadf$q.values),]
taxadf.filt <- taxadf.filt[1:20,]
taxadf.filt <- taxadf.filt[-2, ]

rdaplot <- ggplot()+
  # site scores
    geom_point(aes(scores(allrda, scaling = 2)$sites[,1], 
                   scores(allrda, scaling = 2)$sites[,2], 
                   shape = samples$rotation, 
                   color = as.factor(samples$week)), 
               size = 4)+
  # species scores    
  # geom_point(aes(scores(allrda, scaling = 2)$species[,1], 
  #                    scores(allrda, scaling = 2)$species[,2]), 
  #                alpha = 1/10, color = "gray25", size = 1)+
  # species labels
    # geom_text_repel(aes(x = taxadf.filt[ ,1],
    #               y = taxadf.filt[ ,2],
    #               label = taxadf.filt$spplabels), max.overlaps = 30,
    #           size = 3, color = "gray25", fontface = "italic")+

  # centroid labels
    geom_text(aes(x = scores(allrda, scaling = 2)$centroids[ ,1],
                  y = scores(allrda, scaling = 2)$centroids[ ,2],
                  label = c("sentinel", "bulk", "CS", "CSSwP")), check_overlap = FALSE,
              size = 4, fontface = "bold")+
    xlab("RDA1 - 54% of total variance, p = 0.001")+
    ylab("RDA2 - 2.5% of total variance, p = 0.001")+
    labs(color = "week", shape = "crop\nrotation")+
    theme(legend.text = element_text(size = 12, face = "italic"))+
    theme_minimal()
rdaplot
ggsave(filename = "bac_2020_allRDA_scale2.tiff", plot = rdaplot, device = "tiff", width = 6, height = 4, units = "in")
```

# Sample type RDA
```{r}
bulk <- comm_subset(comm, samples, samp_type = "bulk")
bulksamp <- sample_subset(samples, samp_type = "bulk")

bulkclr <- decostand(bulk+1, method = "clr")
bulkrda <- rda(bulkclr ~ week + rotation, data = bulksamp)
bulkrda
summary(bulkrda)
anova(bulkrda, by = "margin")
anova(bulkrda, by = "axis")

# get taxa to plot
source("r scripts/RDA outlier functions.R")
adaptive <- rdadapt(bulkrda, 2) # finds taxa that contribute most to each axis, returns a list of pvalues

# get taxa that contribute most to each axis
taxadf <- scores(bulkrda, scaling = 2)$species
taxadf <- cbind(taxadf, adaptive)
taxadf$spplabels <- sapply(strsplit(rownames(taxadf), " "), "[[", 1)
# filter to just smbulkest qvalues
taxadf.filt <- taxadf[order(taxadf$q.values),]
taxadf.filt <- taxadf.filt[1:20,]

bulkrdaplot <- ggplot()+
  # site scores
    geom_point(aes(scores(bulkrda, scaling = 2)$sites[,1], 
                   scores(bulkrda, scaling = 2)$sites[,2], 
                   shape = bulksamp$rotation, 
                   color = as.factor(bulksamp$week)), 
               size = 4)+
  # species scores    
  geom_point(aes(scores(bulkrda, scaling = 2)$species[,1], 
                     scores(bulkrda, scaling = 2)$species[,2]), 
                 alpha = 1/10, color = "gray25", size = 1)+
  # species labels
    geom_text_repel(aes(x = taxadf.filt[ ,1],
                  y = taxadf.filt[ ,2],
                  label = taxadf.filt$spplabels), max.overlaps = 30,
              size = 3, color = "gray25", fontface = "italic")+

  # centroid labels
    geom_text(aes(x = scores(bulkrda, scaling = 2)$centroids[ ,1],
                  y = scores(bulkrda, scaling = 2)$centroids[ ,2],
                  label = c("CS", "CSSwP")), check_overlap = FALSE,
              size = 4, fontface = "bold")+
    xlab("RDA1 - 3.9% of total variance, p = 0.007")+
    ylab("RDA2 - 2.2% of total variance, p = 0.3")+
    labs(color = "week", shape = "crop\nrotation")+
    theme(legend.text = element_text(size = 12, face = "italic"))+
    theme_minimal()
bulkrdaplot
ggsave(filename = "bac_2020_bulkRDA_scale2.tiff", plot = bulkrdaplot, device = "tiff", width = 7, height = 5, units = "in")



bag <- comm_subset(comm, samples, samp_type = "bag")
bagsamp <- sample_subset(samples, samp_type = "bag")

bagclr <- decostand(bag+1, method = "clr")
bagrda <- rda(bagclr ~ week + rotation, data = bagsamp)
bagrda
summary(bagrda)
anova(bagrda, by = "margin")
anova(bagrda, by = "axis")

# get taxa to plot
source("r scripts/RDA outlier functions.R")
adaptive <- rdadapt(bagrda, 2) # finds taxa that contribute most to each axis, returns a list of pvalues

# get taxa that contribute most to each axis
taxadf <- scores(bagrda, scaling = 2)$species
taxadf <- cbind(taxadf, adaptive)
taxadf$spplabels <- sapply(strsplit(rownames(taxadf), " "), "[[", 1)
# filter to just smbagest qvalues
taxadf.filt <- taxadf[order(taxadf$q.values),]
taxadf.filt <- taxadf.filt[1:20,]

bagrdaplot <- ggplot()+
  # site scores
    geom_point(aes(scores(bagrda, scaling = 2)$sites[,1], 
                   scores(bagrda, scaling = 2)$sites[,2], 
                   shape = bagsamp$rotation, 
                   color = as.factor(bagsamp$week)), 
               size = 4)+
  # species scores    
  geom_point(aes(scores(bagrda, scaling = 2)$species[,1], 
                     scores(bagrda, scaling = 2)$species[,2]), 
                 alpha = 1/10, color = "gray25", size = 1)+
  # species labels
    geom_text_repel(aes(x = taxadf.filt[ ,1],
                  y = taxadf.filt[ ,2],
                  label = taxadf.filt$spplabels), max.overlaps = 35,
              size = 3, color = "gray25", fontface = "italic")+

  # centroid labels
    geom_text(aes(x = scores(bagrda, scaling = 2)$centroids[ ,1],
                  y = scores(bagrda, scaling = 2)$centroids[ ,2],
                  label = c("CS", "CSSwP")), check_overlap = FALSE,
              size = 4, fontface = "bold")+
    xlab("RDA1 - 25% of total variance, p = 0.001")+
    ylab("RDA2 - 3% of total variance, p = 0.04")+
    labs(color = "week", shape = "crop\nrotation")+
    theme(legend.text = element_text(size = 12, face = "italic"))+
    theme_minimal()
bagrdaplot
ggsave(filename = "bac_2020_bagRDA_scale2.tiff", plot = bagrdaplot, device = "tiff", width = 7, height = 5, units = "in")
```

guild diversity in cropping systems
```{r}
bagdiv <- calculate_diversity(comm_subset(traitcomm, samples, "bag"), sample_subset(samples, "bag"))

bulkdiv <- calculate_diversity(comm_subset(traitcomm, samples, "bulk"), sample_subset(samples, "bulk"))

ggplot(bulkdiv)+
  geom_point(aes(x = week_int, y = rar.hillq1, color = rotation), size = 2)+
  theme_minimal()+
  scale_color_manual(values = c("skyblue", "hotpink"))+
  labs(y = "DNA concentration (ng.uL)", x = "week sampled",title = "bulk soil")
  stat_smooth(aes(x = week_int, y = rar.hillq1), method="lm", se=TRUE, fill=NA,
              formula=y ~ x ,colour="gray")
  
ggplot(bagdiv)+
  geom_point(aes(x = week_int, y = rar.hillq1, color = rotation), size = 3)+
  theme_minimal()+
  scale_color_manual(values = c("skyblue", "hotpink"))+
  labs(y = "DNA concentration (ng.uL)",x = "week sampled", title = "bag soil")
  stat_smooth(aes(x = week_int, y = DNA_ng.ul), method="lm", se=TRUE, fill=NA,
              formula=y ~ exp(-x), colour="gray")
  
t.test(bagdiv$rar.alpha ~ as.numeric(as.factor(bagdiv$rotation)))
t.test(bagdiv$rar.hillq1 ~ as.numeric(as.factor(bagdiv$rotation)))
t.test(bagdiv$rar.shannon ~ as.numeric(as.factor(bagdiv$rotation)))

t.test(bulkdiv$rar.alpha ~ as.numeric(as.factor(bulkdiv$rotation)))
t.test(bulkdiv$rar.hillq1 ~ as.numeric(as.factor(bulkdiv$rotation)))
t.test(bulkdiv$rar.shannon ~ as.numeric(as.factor(bulkdiv$rotation)))

```